---
title: Profiling in R
format: html
---

## Why is My Code So Slow?

* Profiling is a systematic way to examine how much time is spent in different parts of a program
* Useful when optimizing code
* Often code runs fine once, but what if we have to put it in a loop for 1,000 iterations? Is it still fast enough?
* Profiling is better than guessing

## General Principles of Optimization

* Design first, then optimize
* Premature optimization is the root of all evil
* Measure (collect data), don't guess

## Using `system.time()`

* `system.time()` takes an arbitrary R expression as input (can be wrapped in curly braces) and returns the amount of time taken to evaluate that expression.
* It computes the time (in seconds) needed to execute an expression.
  * If there is an error, it returns the time until the error occured.
* Returns and object of class `proc_time`.
  * **user time**: time charged to the CPU(s) for the expression
  * **elapsed time**: "wall clock" time
* Usually, the user time and elapsed time are relatively close, for straight computing tasks
* Elapsed time may be greater than user time if the CPU spends a lot of time waiting around
* Elapsted time may be smaller than the user time if your machine has multiple cores/processors (and is capable of using them)
  * Multi-threaded BLAS libraries (vecLib/Accelerate, ATLAS, ACML, MKL)
  * Parallel processing via the `parallel` package

```{r}
# Elapsed time > user time
system.time(readLines("https://www.google.com"))

# Elapsed time < user time
hilbert <- function(n) {
    i <- 1:n
    1 / outer(i - 1, i, "+")
}
x <- hilbert(1000)
system.time(svd(x))
```

### Timing Longer Expressions

```{r}
system.time({
    n <- 1000
    r <- numeric(n)

    for (i in 1:n) {
        x <- rnorm(x)
        r[i] <- mean(x)
    }
})
```

## The R Profiler

The `Rprof()` function starts the profiler in R. (R must be compiled with profiler support). `Rprof()` keeps track of the function call stack at regularly sampled intervals and tabulates how much time is spent on each function. Default sampling interval is 0.02 seconds.

> DO NOT use `system.time()` and `Rprof` together!

### Using `summaryRprof`

The `summaryRprof()` function tabulates the R profiler output and calculates how much time is spent on which function, summarizing it.

There are two methods to normalize the data:

* `by.total`, which divides the time spent on each by the total runtime
* `by.self`, which does the same but first subtracts out time the time spent in function above the call stack


