---
title: Simulation in R
format: html
---

## Random Numbers Generation

Functions for proability distributions in R:

* `rnorm`: Generate random normal variates with a given mean and standard deviation
* `dnorm`: Evaluate the normal probability density (with a given mean, sd) at a point (or vector of points)
* `pnorm`: Evaluate the cumulative distribution function for a normal distribution
* `rpois`: Generate random Poisson variates with a given rate

Probability distribution functions usually have four classes associated with them. The functions are prefixed with:

* `d` for density
* `r` for random number generation
* `p` for cumulative distribution
* `q` for quantile function

Working with normal distributions require these for functions:

```r
dnorm(x, mean = 0, sd = 1, log = FALSE)
pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
rnorm(n, mean = 0, sd = 1)
```

If $\Phi$ is the cumulative distribution function (CDF) of the standard normal distribution, then `pnorm(q)` $= \Phi(q)$ and `qnorm(p)` $= \Phi^{-1}(p)$, where $\Phi^{-1}$ denotes the quantile function (the inverse CDF).

For example, to generate ten random numbers, we can call:

```{r}
x <- rnorm(10)
x
summary(x)
```

This generates 10 normal numbers with $mean = 0$ and $sd = 1$.

If we want to generate a vector with $mean = 20$ and $sd = 2$, we can call:

```{r}
y <- rnorm(10, 20, 2)
y
summary(y)
```

When generating random numbers, it's important to set the random number seed with `set.seed` to ensure reproducibility.

For example:

```{r}
set.seed(1)
rnorm(5)
rnorm(5)
set.seed(1)
rnorm(5)
```

We can see the first two calls to `rnorm()` result in different vectors of random numbers. However, after calling `set.seed(1)` and generating a set of five random numbers again, we get the same vector as the first call to `rnorm()`.

### Generating Poisson Data

We can also generate random variables for other probability distributions. For example, here we'll generate ten Poisson random variables with a rate of 1, ten with a rate of 2, and ten with a rate of 20.

```{r}
rpois(10, 1)
rpois(10, 2)
rpois(10, 20)
```

We can also evaluate the cumulative distribution function for the Poisson distribution.

```{r}
ppois(2, 2)  # Pr(x <= 2)
ppois(4, 2)  # Pr(x <= 4)
ppois(6, 2)  # Pr(x <= 6)
```

In the example above, we calculate the probabilities of a Poisson random variable to be less than or equal to two, four, and six (respectively) if its rate if two.

## Generating Random Numbers From a Linear Model

Suppose we want to simulate from the following linear model:

$$
y = \beta_0 + \beta_1 x + \varepsilon
$$

$$
\text{where} \quad \varepsilon \sim \mathcal{N}(0,\, 2^2). \quad \text{Assume} \quad x \sim \mathcal{N}(0,\, 1^2), \quad \beta_0 = 0.5 \quad \text{and} \quad \beta_1 = 2
$$

The linear model above has a single predictor, $x$ and a random noise, $\varepsilon$, which has a normal distribution and a standard deviation of 2. The output will be generated by the coefficients $\beta_0$ and a slope of $\beta_1$, and we have to assume that $\beta_0 = 0.5$ and $\beta_1 = 2$.

So the question is *How do we simulate from this model*?


```{r}
# We start by setting the `seed` to `20`
set.seed(20)

# Then, we generate the predictor, which has a standard normal distribution
x <- rnorm(100)

# We, then, generate ε, which has a standard normal distribution, a mean of zero and sd of 2
e <- rnorm(100, 0, 2)

# Then, we simulate the adding operation
y <- 0.5 + 2 * x + e

summary(y)
plot(x, y)
```

Looking at the plot we can see that `x` and `y` have a linear relationship, as expected.

Now, to make a variation of it, let's say `x` is a binary random variable, maybe representing a gender or a treatment versus control scenario. For that, we can generate binary data using a binomial distribution via the `rbinom()` function.

```{r}
set.seed(10)
x <- rbinom(100, 1, 0.5)
e <- rnorm(100, 0, 2)
y <- 0.5 + 2 * x + e
summary(y)
plot(x, y)
```

As expected, we see a different graph now, because the `x` variable is binary. However, `y` is still continous, it's normal. So, we see a linear trend one more time.

Now, let's suppose we want to simulate a generalized linear model from a slightly more complicated model, perhaps with a Poisson distribution.

So, we want to simulate from a Poisson model where:

$$
Y \sim \text{Poisson}(\mu)
$$

$$
\log(\mu) = \beta_0 + \beta_1 x, \quad \beta_0 = 0.5, \quad \beta_1 = 0.3
$$

We'll need the `rpois` function for this:

```{r}
set.seed(1)
x <- rnorm(100)
log.mu <- 0.5 + 0.3 * x
y <- rpois(100, exp(log.mu))

summary(y)
plot(x, y)
```

### Random Sampling

The `sample()` function draws randomly from a specified set of (scalar) objects, allowing you to sample from arbitrary distributions.

```{r}
set.seed(1)
sample(1:10, 4)
sample(1:10, 4)
sample(letters, 5)
sample(1:10)  # Permutation
sample(1:10)
sample(1:10, replace = TRUE)  # Sample with replacement
```
