---
title: Programming Assignment 3
format: html
---

The data for this assignment come from the [Hospital Compare](http://hospitalcompare.hhs.gov) web site run by the U.S. Department of Health and Human Services. The purpose of the web site is to provide data and information about the quality of care at over 4,000 Medicare-certified hospitals in the U.S. This dataset essentially covers all major U.S. hospitals. This dataset is used for a variety of purposes, including determining whether hospitals should be fined for not providing high quality care to patients.

The Hospital Compare web site contains a lot of data and we will only look at a small subset for this assignment.
* `outcome-of-care-measures.csv`: Contains information about 30-day mortality and readmission rates for heart attacks, heart failure, and pneumonia for over 4,000 hospitals.
* `hospital-data.csv`: Contains information about each hospital.

## 1. Plot the 30-day mortality rates for heart attack

Read the outcome data into R via the `read.csv` function and look at the first few rows.

```{r}
outcome <- read.csv("outcome-of-care-measures.csv", colClass = "character")
# head(outcome)
# summary(outcome)
outcome[, 11] <- as.numeric(outcome[, 11])  # Coerce character column to numeric values
hist(outcome[, 11])
```

## 2. Finding the best hospital in a state

Write a function called `FindBestHospital` that take two arguments: the 2-character abbreviated name of a state and an outcome name. The function reads the `outcome-of-care-measures.csv` file and returns a character vector with the name of the hospital that has the best (i.e. lowest) 30-day mortality for the specified outcome in that state. The hospital name is the name provided in the `Hospital.Name` variable. The outcomes can be one of “heart attack”, “heart failure”, or “pneumonia”. Hospitals that do not have data on a particular outcome should be excluded from the set of hospitals when deciding the rankings.

**Handling ties**. If there is a tie for the best hospital for a given outcome, then the hospital names should be sorted in alphabetical order and the first hospital in that set should be chosen (i.e. if hospitals “b”, “c”, and “f” are tied for best, then hospital “b” should be returned).

The function should check the validity of its arguments. If an invalid `state` value is passed to `FindBestHospital`, the function should throw an error via the `stop` function with the exact message “invalid state”. If an invalid `outcome` value is passed to best, the function should throw an error via the stop function with the exact message “invalid outcome”.

### Solution

```{r}
library(dplyr)

FindBestHospital <- function(state, outcome) {

    hospitals_data <- read.csv("outcome-of-care-measures.csv", colClass = "character")
    STATES <- unique(hospitals_data$State)
    OUTCOMES <- c("heart attack", "heart failure", "pneumonia")

    if (!is.element(state, STATES)) stop("invalid state")
    if (!is.element(outcome, OUTCOMES)) stop("invalid outcome")

    COL_MAP <- list("heart attack" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack",
                    "heart failure" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure",
                    "pneumonia" = "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia")

    best_hospital <- hospitals_data %>%
        filter(State == state) %>%
        mutate(rate = as.numeric(.data[[COL_MAP[[outcome]]]])) %>%
        filter(!is.na(rate)) %>%
        arrange(rate, Hospital.Name) %>%
        slice(1) %>%
        pull(Hospital.Name)

    best_hospital
}
```

The following test scenarios have been validated in the `FindBestHospital.R` file.

```r
FindBestHospital("TX", "heart attack")
# [1] "CYPRESS FAIRBANKS MEDICAL CENTER"

FindBestHospital("TX", "heart failure")
# [1] "FORT DUNCAN MEDICAL CENTER"

FindBestHospital("MD", "heart attack")
# [1] "JOHNS HOPKINS HOSPITAL, THE"

FindBestHospital("MD", "pneumonia")
# [1] "GREATER BALTIMORE MEDICAL CENTER"

FindBestHospital("BB", "heart attack")
# Error in FindBestHospital("BB", "heart attack") : invalid state

FindBestHospital("NY", "hert attack")
# Error in FindBestHospital("NY", "hert attack") : invalid outcome
```

## 3. Ranking hospitals by outcome in a state

Write a function called `RankHospital` that takes three arguments: the 2-character abbreviated name of a state (`state`), an outcome (`outcome`), and the ranking of a hospital in that state for that outcome (`num`). The function reads the `outcome-of-care-measures.csv` file and returns a character vector with the name of the hospital that has the ranking specified by the `num` argument. For example, the call to

```r
RankHospital("MD", "heart failure", 5)
```

would return a character vector containing the name of the hospital with the 5th lowest 30-day death rate for heart failure. The `num` argument can take values “best”, “worst”, or an integer indicating the ranking (smaller numbers are better). If the number given by `num` is larger than the number of hospitals in that state, then the function should return `NA`. Hospitals that do not have data on a particular outcome should be excluded from the set of hospitals when deciding the rankings.

**Handling ties**. It may occur that multiple hospitals have the same 30-day mortality rate for a given cause of death. In those cases, ties should be broken by using the hospital name. For example, in Texas (“TX”), the hospitals with lowest 30-day mortality rate for heart failure are shown here.

```r
head(texas)

#                         Hospital.Name Rate Rank
# 3935       FORT DUNCAN MEDICAL CENTER  8.1    1
# 4085  TOMBALL REGIONAL MEDICAL CENTER  8.5    2
# 4103 CYPRESS FAIRBANKS MEDICAL CENTER  8.7    3
# 3954           DETAR HOSPITAL NAVARRO  8.7    4
# 4010           METHODIST HOSPITAL,THE  8.8    5
# 3962  MISSION REGIONAL MEDICAL CENTER  8.8    6
```

Note that Cypress Fairbanks Medical Center and Detar Hospital Navarro both have the same 30-day rate (8.7). However, because Cypress comes before Detar alphabetically, Cypress is ranked number 3 in this scheme and Detar is ranked number 4. One can use the order function to sort multiple vectors in this manner (i.e. where one vector is used to break ties in another vector).

The function should check the validity of its arguments. If an invalid `state` value is passed to `rankhospital`, the function should throw an error via the `stop` function with the exact message “invalid state”. If an invalid `outcome` value is passed to `rankhospital`, the function should throw an error via the `stop` function with the exact message “invalid outcome”.

### Solution

```{r}
RankHospital <- function(state, outcome, num) {
    hospitals_data <- read.csv("outcome-of-care-measures.csv", colClass = "character")

    STATES <- unique(hospitals_data$State)
    OUTCOMES <- c("heart attack", "heart failure", "pneumonia")

    if (!is.element(state, STATES)) stop("invalid state")
    if (!is.element(outcome, OUTCOMES)) stop("invalid outcome")

    COL_MAP <- list(
        "heart attack" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack",
        "heart failure" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure",
        "pneumonia"     = "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia"
    )

    ranked <- hospitals_data %>%
        filter(State == state) %>%
        mutate(Rate = as.numeric(.data[[COL_MAP[[outcome]]]])) %>%
        filter(!is.na(Rate)) %>%
        arrange(Rate, Hospital.Name)

    idx <- if (num == "best") 1
           else if (num == "worst") nrow(ranked)
           else num

    if (idx > nrow(ranked)) return(NA)

    ranked %>% slice(idx) %>% pull(Hospital.Name)
}
```

### Validation

```{r}
RankHospital("TX", "heart failure", 4)  # [1] "DETAR HOSPITAL NAVARRO"
RankHospital("MD", "heart attack", "worst")  # [1] "HARFORD MEMORIAL HOSPITAL"
RankHospital("MN", "heart attack", 5000)  # [1] NA
```

## 4. Ranking hospitals in all states

Write a function called `RankAllHospitals` that takes two arguments: an outcome name (`outcome`) and a hospital ranking (`num`). The function reads the `outcome-of-care-measures.csv` file and returns a 2-column data frame containing the hospital in each state that has the ranking specified in `num`. For example, the function call `RankAllHospitals("heart attack", "best")` would return a data frame containing the names of the hospitals that are the best in their respective states for 30-day heart attack death rates. The function should return a value for every state (some may be `NA`). The first column in the data frame is named `hospital`, which contains the hospital name, and the second column is named `state`, which contains the 2-character abbreviation for the state name. Hospitals that do not have data on a particular outcome should be excluded from the set of hospitals when deciding the rankings.

**Handling ties**. The `RankAllHospitals` function should handle ties in the 30-day mortality rates in the same way that the `RankHospital` function handles ties.

**NOTE**: For the purpose of this part of the assignment (and for efficiency), your function should NOT call the `RankHospital` function from the previous section.

The function should check the validity of its arguments. If an invalid `outcome` value is passed to `RankAllHospitals`, the function should throw an error via the `stop` function with the exact message “invalid outcome”. The `num` variable can take values “best”, “worst”, or an integer indicating the ranking (smaller numbers are better). If the number given by `num` is larger than the number of hospitals in that state, then the function should return `NA`.

### Solution

```{r}
RankAllHospitals <- function(outcome, num) {
    hospitals_data <- read.csv("outcome-of-care-measures.csv", colClass = "character")

    OUTCOMES <- c("heart attack", "heart failure", "pneumonia")

    if (!is.element(outcome, OUTCOMES)) stop("invalid outcome")

    COL_MAP <- list(
        "heart attack" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Attack",
        "heart failure" = "Hospital.30.Day.Death..Mortality..Rates.from.Heart.Failure",
        "pneumonia"     = "Hospital.30.Day.Death..Mortality..Rates.from.Pneumonia"
    )

    hospitals_data %>%
        mutate(Rate = as.numeric(.data[[COL_MAP[[outcome]]]])) %>%
        filter(!is.na(Rate)) %>%
        arrange(Rate, Hospital.Name) %>%
        group_by(State) %>%
        summarise(hospital = {
            idx <- if (num == "best") 1 else if (num == "worst") n() else as.integer(num)
            if (idx > n()) NA_character_ else Hospital.Name[idx]
        }) %>%
        select(hospital, state = State)
}
```

### Validations

```{r}
head(RankAllHospitals("heart attack", 20), 10)
tail(RankAllHospitals("pneumonia", "worst"), 3)
```
