---
title: Air Pollution Programming Assignment
format: html
---

In this programming assignment, I was tasked with writing three functions
that were meant to interact with the dataset that accompanies this assignment.
The dataset is contained in the `specdata` directory.

## Data

The `specdata` directory contains 332 CSV files containing pollution monitoring
data for fine particulate matter (PM) air pollution at 332 locations in the US.
Each file contains data from a single monitor and the ID number
for each monitor is contained in the file name. For example,
data for monitor 200 is contained in the file `200.csv`.
Each file contains three variables:

- Date: the date of the observation in `YYYY-MM-DD` format (year-month-day)
- sulfate: the level of sulfate PM in the air on that date
  (measured in micrograms per cubic meter)
- nitrate: the level of nitrate PM in the air on that date
  (measured in micrograms per cubic meter)

> In each file there are many days when either sulfate or nitrate
> (or both) are missing (coded as NA). This is common with air pollution
> monitoring data in the United States.

## Setup

To run Quarto notebooks, we need to install the rmarkdown, tidyverse,
and palmerpenguins packages:

```{r}
install.packages("rmarkdown", repos = "https://cloud.r-project.org")
install.packages("tidyverse", repos = "https://cloud.r-project.org")
install.packages("palmerpenguins", repos = "https://cloud.r-project.org")

library(rmarkdown)
library(tidyverse)
library(palmerpenguins)
```

## Part 1 — Pollutant Mean Analysis Across Monitoring Stations

Write a function named `pollutantmean` that calculates the mean of a pollutant
(sulfate or nitrate) across a specified list of monitors. The `pollutantmean`
function takes three arguments: `directory`, `pollutant`, and `id`.
Given a vector monitor ID numbers, `pollutantmean` reads that monitors'
particulate matter data from the directory specified in the `directory`
argument and returns the mean of the pollutant across all the monitors,
ignoring any missing values coded as NA.

```{r}
pollutantmean <- function(directory, pollutant, id = 1:332) {
    # Create an empty numeric vector
    # This will store pollutant values from all monitors
    pollutant_values <- numeric()

    # Loop through each monitor id
    for (i in id) {

        # Build the file name using 3-digit formatting
        # Example: 1 -> "001.csv"
        file_name <- sprintf("%03d.csv", i)

        # Create the complete file path
        # Example: "specdata/001.csv"
        file_path <- file.path(directory, file_name)

        # Read the CSV file into a data frame
        data <- read.csv(file_path)

        # Extract the selected pollutant column
        # pollutant can be "sulfate" or "nitrate"
        pollutant_column <- data[[pollutant]]

        # Append pollutant values to the main vector
        pollutant_values <- c(
            pollutant_values,
            pollutant_column
        )
    }

    # Calculate the mean while ignoring missing values (NA)
    mean(pollutant_values, na.rm = TRUE)
}
```

### Testing the `pollutantmean` function

```{r}
pollutantmean("specdata", "sulfate", 1:10)   # Prints 4.064128
pollutantmean("specdata", "nitrate", 70:72)  # Prints 1.70604
pollutantmean("specdata", "nitrate", 23)     # Prints 1.280833
```

### Notes and Observations (Part 1)

This exercise is more important than it initially appears because it introduces
several foundational concepts used constantly in real-world Data Science
and Data Engineering workflows.

At a high level, we are building a small-scale data ingestion and aggregation
pipeline.

The exercise teaches us how to:

* Read data from multiple files
* Dynamically generate file paths
* Iterate through datasets programmatically
* Extract only relevant variables
* Handle missing data (NA)
* Aggregate values across distributed datasets
* Compute summary statistics

Conceptually, this is a simplified version of what happens in production
analytics systems.

For example:

* Each CSV file represents a separate data source or sensor
* The `specdata` directory behaves like a distributed dataset
* Our function acts as a data processing job
* `mean()` becomes a statistical aggregation step

This mirrors real systems where:

* IoT sensors produce separate files
* Logs are partitioned by date or server
* Data lakes store datasets across many objects/files

The most important data science concepts hidden inside this exercise are:

1. Data ingestion
    Loading raw data into memory for analysis.
2. Data wrangling
    Extracting only needed columns and combining datasets.
3. Missing data handling
    Real-world datasets are incomplete. Ignoring NA correctly is critical.
4. Functional abstraction
    You are creating reusable analysis logic instead of writing repetitive
    code.
5. Statistical aggregation
    The mean is one of the simplest summary statistics, but the same pattern
    scales to:

    * medians
    * correlations
    * regressions
    * machine learning feature generation

Architecturally, this exercise is also teaching us an important mindset shift:

Instead of thinking:

> “I manually analyze one file”

we begin thinking:

> “I design a system that analyzes arbitrary datasets automatically.”

## Part 2 — Complete Case Analysis Across Monitoring Stations

Write a function that reads a directory full of files and reports the number
of completely observed cases in each data file. The function should return
a data frame where the first column is the name of the file and the second
column is the number of complete cases.

```{r}
complete <- function(directory, id = 1:332) {
    # Create an empty data frame to store results
    # id   -> monitor ID
    # nobs -> number of complete observations
    result <- data.frame(
        id = integer(),
        nobs = integer()
    )

    # Loop through each monitor ID
    for (i in id) {
      
        # Create file name with leading zeros
        # Example: 1 -> "001.csv"
        file_name <- sprintf("%03d.csv", i)

        # Build the complete file path
        file_path <- file.path(directory, file_name)

        # Read the CSV file into a data frame
        data <- read.csv(file_path)

        # Identify rows with no missing values
        # complete.cases() returns TRUE/FALSE for each row
        complete_rows <- complete.cases(data)

        # Count how many rows are completely observed
        n_complete <- sum(complete_rows)

        # Add monitor result to the output data frame
        result <- rbind(
            result,
            data.frame(
                id = i,
                nobs = n_complete
            )
        )
    }

    # Return final results table
    result
}
```


### Testing the `complete` function

```{r}
complete("specdata", 1)
complete("specdata", c(2, 4, 8, 10, 12))
complete("specdata", 30:25)
complete("specdata", 3)
```

### Notes and Observations (Part 2)

This exercise is fundamentally about data quality assessment,
which is one of the most important and underestimated parts of Data Science.

The function is not really about counting rows. It is about answering a much
deeper question:

> “How reliable is the data source?”

In real-world analytics and ML systems, missing data is everywhere:

* sensor failures
* incomplete forms
* broken pipelines
* API outages
* corrupted records
* human error

Before building models or generating insights, we must first determine
whether the data is trustworthy enough to use.

This exercise teaches several critical concepts.

First, it introduces the idea of complete observations:

* A complete case is a row with no missing values
* In this dataset, a row is only complete if all variables exist (Date, sulfate,
  nitrate)

This matters because many statistical methods and ML algorithms assume
complete data.

Second, it teaches data profiling. Our function is effectively profiling
each monitor:

* Which monitors have reliable data?
* Which monitors have sparse or incomplete observations?
* Which sources should be trusted more?

This is exactly what happens in production data systems.

For example:

* A data engineer checks pipeline completeness
* A machine learning engineer validates training data quality
* A data scientist evaluates whether missingness introduces bias

Third, this exercise teaches an important analytical mindset:
we are not just consuming data—we are auditing it. That shift is critical.

Architecturally, the function also teaches:

* Iteration over distributed datasets
* Aggregation into summary tables
* Boolean masking (complete.cases)
* Building reusable data quality utilities

The output is essentially a miniature data quality report.

In larger systems, this same pattern scales into:

* pipeline observability
* data validation frameworks
* monitoring dashboards
* feature quality metrics
* ML data contracts

## Part 3 — Correlation Analysis Between Sulfate and Nitrate Levels

Write a function that takes a directory of data files and a threshold
for complete cases and calculates the correlation between sulfate and nitrate
for monitor locations where the number of completely observed cases
(on all variables) is greater than the threshold. The function should return
a vector of correlations for the monitors that meet the threshold requirement.
If no monitors meet the threshold requirement, then the function should return
a numeric vector of length `0`.

```{r}
corr <- function(directory, threshold = 0) {

    # Create an empty numeric vector to store correlations
    correlations <- numeric()

    # Get all CSV file names from the directory
    files <- list.files(directory, full.names = TRUE)

    # Loop through each file
    for (file in files) {

        # Read current CSV file
        data <- read.csv(file)

        # Identify rows with complete observations        
        # (no missing values in any column)
        complete_rows <- complete.cases(data)

        # Count how many complete observations exist
        n_complete <- sum(complete_rows)

        # Only calculate correlation if the monitor
        # has more complete cases than the threshold
        if (n_complete > threshold) {

            # Subset data to only complete rows
            complete_data <- data[complete_rows, ]

            # Calculate correlation between sulfate and nitrate
            correlation_value <- cor(
                complete_data$sulfate,
                complete_data$nitrate
            )

            # Append correlation result to vector
            correlations <- c(
                correlations,
                correlation_value
            )
        }
    }

    # Return vector of correlations
    correlations
}
```

### Testing the `corr` function

```{r}
cr <- corr("specdata", 150)
head(cr)
summary(cr)

cr <- corr("specdata", 400)
head(cr)
summary(cr)

cr <- corr("specdata", 5000)
summary(cr)
length(cr)

cr <- corr("specdata")
summary(cr)
length(cr)

cr <- corr("specdata")
cr <- sort(cr)
set.seed(868)
out <- round(cr[sample(length(cr), 5)], 4)
print(out)
```

### Notes and Observations (Part 3)

This exercise is extremely important because it introduces one of the most
fundamental concepts in Statistics, Data Science, and Machine Learning:

> understanding relationships between variables.

The previous exercises focused on:

* aggregation (mean)
* data quality (complete cases)

This third exercise moves into:

* statistical relationships
* feature interaction analysis
* exploratory data analysis (EDA)

That is a major conceptual step forward.

The `corr()` function measures how strongly two pollutants (sulfate and nitrate)
move together.

A positive correlation means:

* when sulfate increases, nitrate tends to increase

A negative correlation means:

* when sulfate increases, nitrate tends to decrease

A near-zero correlation means:

* little or no linear relationship

This is foundational because almost every advanced analytical
workflow begins with relationship analysis.

Examples:

* In finance → correlation between stocks
* In healthcare → correlation between biomarkers
* In cybersecurity → correlation between system events
* In recommender systems → correlation between user behaviors
* In ML → correlation between features and targets

This assignment also introduces an extremely important real-world concept:

> not all data sources are equally reliable.

That is why the threshold parameter exists.

Instead of blindly computing correlations on every monitor, we first verify:

* does this monitor contain enough complete observations?

This is a primitive form of data validation and statistical reliability filtering.

A monitor with five complete observations should not be trusted as much as one
with 1000 complete observations.

So the exercise teaches us:

* statistical computation
* data filtering
* quality thresholds
* conditional processing
* reproducible analysis pipelines

From an architectural perspective, the function behaves like a distributed
analytics job:

1. Iterate through multiple data sources
2. Validate data quality
3. Compute statistical metrics
4. Aggregate results

That pattern appears everywhere in modern data systems.

The most important data science mindset introduced here is this:

Before interpreting a statistical relationship, validate whether the underlying
data is sufficient and trustworthy.

That principle becomes critical later in:

* ML feature selection
* causal inference
* A/B testing
* anomaly detection
* predictive modeling

