---
title: Functional Loops in R
format: html
---

## The Mental Model

In Python you'd write `for item in collection: ...`. R treats functions as first-class citizens, so instead of *loops*, you *apply* a function to every element of a structure. Think of it like an assembly line: the collection is the conveyor belt, your function is the worker at each station, and the output is whatever the worker produces — a vector, a list, a data frame, or nothing (side effects only).

The family splits into three lineages: **base R apply family**, **base R higher-order functions** (`Reduce`, `Filter`, `Map`), and `purrr` (the tidyverse rewrite with type safety).

## Base R — the `apply` family

### `apply()` — for matrices and arrays

The grandfather. Works only on 2D+ structures.

```{r}
# MARGIN = 1 → apply fn to each ROW
# MARGIN = 2 → apply fn to each COLUMN
m <- matrix(1:12, nrow = 3)

apply(m, 1, sum)
apply(m, 2, mean)
```

Analogy: `apply(m, 1, fn)` is like slicing a loaf of bread horizontally — you get every row as a slice, then you do something to each slice.

### `lapply()` — list in, list out. Always.

The workhorse. Takes any vector or list, applies a function to each element, and **always returns a list** — no surprises, no coercion.

```{r}
nums <- list(a = 1:5, b = 6:10, c = 11:15)

lapply(nums, mean)
```

Use `lapply` when you want to preserve structure, or when outputs are heterogeneous (different types per element).

### `sapply()` — `lapply` with automatic simplification

Same contract as `lapply`, but tries to simplify the result to a vector or matrix. The tradeoff: convenience vs predictability.

```{r}
sapply(nums, mean)
```

Returns a named numeric vector, not a list.

**The trap**: if your function returns vectors of different lengths across elements, `sapply` falls back to a list silently — your type guarantee evaporates. That's why `vapply` exists.

### `vapply()` — `sapply` with a type contract

You specify *exactly* what output type and shape to expect. If the function deviates, it errors immediately rather than silently returning a list. This is the production-safe version.

```{r}
# Third arg: the expected return type template (numeric(1) = one double per call)
vapply(nums, mean, numeric(1))
```

**Rule of thumb**: use `sapply` interactively for exploration, `vapply` in any code that will run unattended (scripts, pipelines, package internals).

### `tapply()` — apply by group (like a GROUP BY)

Splits a vector into groups defined by a factor, applies a function to each group. Think SQL `GROUP BY` + aggregate.

```{r}
scores <- c(85, 92, 78, 90, 88, 76)
groups <- c("A", "B", "A", "B", "A", "B")

tapply(scores, groups, mean)
```

Internally, `tapply` is `split()` + `lapply()` fused into one call — which brings us to the utility functions below.

#### `mapply()` — multivariate `sapply`

When your function takes *multiple* arguments and you want to vectorize over all of them in parallel.

```{r}
mapply(function(x, y) x ^ y,
       x = c(2, 3, 4),
       y = c(3, 2, 1))
# [1] 8 9 4     ← 2^3, 3^2, 4^1
```

#### `rapply()` — recursive apply

Operates on arbitrarily nested lists, walking the entire tree. Useful for deeply nested JSON-like structures.

```{r}
nested <- list(a = list(x = 1:3, y = 4:6), b = list(x = 7:9))

rapply(nested, mean, how = "unlist")

# how = "replace" → preserves the original list structure
rapply(nested, mean, how = "replace")
```

## Base R — higher-order functions

These are the functional programming primitives — clean, composable, and often underused.

### `Reduce()` — **fold** / **accumulate**

Collapses a list or vector into a single value by repeatedly applying a binary function. This is the classic **fold** from FP.

```{r}
# Sum via fold: ((1 + 2) + 3) + 4 + 5
Reduce("+", 1:5)

# Accumulate intermediate results (like cumsum)
Reduce("+", 1:5, accumulate = TRUE)
# [1]  1  3  6 10 15

# Right fold (from the right side)
Reduce("+", 1:5, right = TRUE, accumulate = TRUE)
```

**Analogy**: imagine a snowball rolling down a hill, picking up more snow at each step. `Reduce` is the hill; each element is a patch of snow.

### `Filter()` — keep elements matching a predicate

```{r}
evens <- Filter(function(x) x %% 2 == 0, 1:10)
```

Companion functions in the same family:

| Function         | What it does                                      |
|------------------|---------------------------------------------------|
| `Filter(f, x)`   | keep elements where `f(x)` is `TRUE`              |
| `Find(f, x)`     | return the *first* element where `f(x)` is `TRUE` |
| `Position(f, x)` | return the *index* of the first match             |
| `Negate(f)`      | return the logical complement of predicate `f`    |

```{r}
# Negate is incredibly useful
not_even <- Negate(function(x) x %% 2 == 0)
Filter(not_even, 1:10)
```

### `Map()` — parallel `lapply` over multiple inputs

The base R equivalent of `mapply` but always returns a list (no simplification).

```{r}
Map(function(x, y) x + y, 1:3, 4:6)
```

## Utility functions

### `split()` — divide and conquer

`split(x, f)` divides `x` into a named list of sub-objects, grouped by factor `f`. By itself it's just a slicer, but paired with `lapply` it becomes one of R's most powerful patterns.

```{r}
df <- data.frame(
    region = c("N", "S", "N", "S", "N"),
    sales = c(100, 200, 150, 250, 120)
)

# Split df by region → list of two mini data frames
groups <- split(df, df$region)

# Now apply any function to each mini data frame
lapply(groups, function(g) mean(g$sales))
```

This pattern — `split` → `lapply` → `do.call(rbind, ...)` — is the base R equivalent of `dplyr::group_by() |> summarise()`.

### `do.call()` — call a function with a list as its argument list

When your arguments live in a list, `do.call` unpacks them and passes each one as a named argument.

```{r}
args <- list(1:10, na.rm = TRUE)
do.call(mean, args)

# Classic use: rbind a list of data frames into one
frames <- list(data.frame(x = 1:2), data.frame(x = 3:4))
do.call(rbind, frames)
```

### `purrr` — the tidyverse upgrade

`purrr` reimplements the `apply` family with consistent naming, strict type safety, and better composability. If you're in a tidyverse pipeline, prefer `purrr`.

**Naming convention**: `map_{type}()`

| Function          | Output type                  |
|-------------------|------------------------------|
| `map(.x, .f)`     | list (like `lapply`)         |
| `map_lgl(.x, .f)` | logical vector               |
| `map_int(.x, .f)` | integer vector               |
| `map_dbl(.x, .f)` | double vector                |
| `map_chr(.x, .f)` | character vector             |
| `map_df(.x, .f)`  | data frame (via `bind_rows`) |

```{r}
library(purrr)

map_dbl(list(a = 1:5, b = 6:10), mean)
```

Unlike `sapply`, `map_dbl` will error loudly if any element doesn't return a single double. That's the guarantee.

### Multi-input: `map2() `and `pmap()`

```{r}
# Two parallel inputs
map2_dbl(c(1, 2, 3), c(10, 20, 30), ~ .x * .y)

# Arbitrary number of inputs (pass a list of lists)
params <- list(mean = c(0, 5, 10), sd = c(1, 2, 3))
pmap(params, rnorm, n = 3)    # generates 3 normal samples per row of params
```

#### Side effects only: `walk()`

When you don't need the return value — writing files, printing, updating a database. Returns `.x` invisibly so pipelines stay composable.

```{r}
walk(list("a.csv", "b.csv", "c.csv"), function(f) {
    cat("Processing: ", f, "\n")
})
```

## Decision Cheat Sheet

| Situation                            | Use                                   |
|--------------------------------------|---------------------------------------|
| Apply fn to matrix rows/cols         | `apply(m, 1_or_2, fn)`                |
| List in, list out, safe              | `lapply(x, fn)`                       |
| List in, simplified out, interactive | `sapply(x, fn)`                       |
| List in, simplified out, production  | `vapply(x, fn, type_template)`        |
| Group-by aggregate                   | `tapply(x, group, fn)`                |
| Multiple parallel inputs             | `mapply(fn, x, y)` or `Map(fn, x, y)` |
| Nested list traversal                | `rapply(x, fn, how=)`                 |
| Fold to single value                 | `Reduce(fn, x)`                       |
| Keep matching elements               | `Filter(predicate, x)`                |
| Group data, then apply               | `split(x, f)` + `lapply(...)`         |
| Bind arg list to a fn                | `do.call(fn, arg_list)`               |
| Tidyverse pipeline, type-strict      | `map_dbl/chr/df(...)`                 |
| Two inputs, tidyverse                | `map2(x, y, fn)`                      |
| N inputs, tidyverse                  | `pmap(list_of_args, fn)`              |
| Side effects only                    | `walk(x, fn)`                         |

## Examples of `map_df`, `map2`, and `pmap`

### `map_df()`

**Scenario**: Summarise each group of exam scores into a one-row data frame, then bind all rows

* Input: list of vectors
* Output: data frame

```{r}
# Each element of the list is a named score vector
scores <- list(
    math    = c(85, 90, 78, 92, 88),
    science = c(70, 75, 80, 85, 72),
    english = c(91, 88, 94, 79, 85)
)

map_df(scores, function(x) {
    data.frame(
        mean   = round(mean(x), 1),
        sd     = round(sd(x), 1),
        top    = max(x),
        passed = sum(x >= 80)
    )
}, .id = "subject")    # .id names the source element
```

Equivalent with tilde shorthand (purrr formula style):

```{r}
map_df(scores,
    ~ data.frame(mean = round(mean(.x), 1),
                 passed = sum(.x >= 80)),
    .id = "subject"
)
```

`.x` refers to the current element — the tilde creates an anonymous function on the fly.

### `map2()` / `map2_dbl()`

**Scenario**: Calculate compound interest for each principal + rate pair.

* Input: Two parallel vectors
* Output: List or typed vector

```{r}
principals <- c(1000, 2500, 5000, 10000)
rates      <- c(0.03, 0.05, 0.07, 0.10)

# map2 → list; map2_dbl → double vector (type-safe)
map2_dbl(principals, rates, function(p, r) {
    round(p * (1 + r)^10, 2)
})

# tilde shorthand: .x = first input, .y = second
map2_dbl(principals, rates, ~ round(.x * (1 + .y)^10, 2))
```

#### `map2` vs `map2_dbl` — the type contract

```{r}
# map2 → always a list, never errors on mixed types
map2(c(1, 2, 3), c("a", "b", "c"), ~ paste(.x, .y))

# map2_chr → typed, errors if fn returns non-character
map2_chr(c(1, 2, 3), c("a", "b", "c"), ~ paste(.x, .y))
```

### `pmap()` / `pmap_dbl()`

**Scenario**: Generate N random normal samples, each with its own n, mean, and sd.

* Input: List of N parallel vectors
* Output: List or typed vector
* N >= 3 inputs

```{r}
# A named list — each row is one function call's arguments
params <- list(
    n    = c(5,   3,   4  ),
    mean = c(0,   10,  -5 ),
    sd   = c(1,   2,   0.5)
)

# pmap passes each column as a named argument to rnorm()
pmap(params, rnorm)

# With a data frame (each row = one call)
df_params <- data.frame(
  n    = c(5, 3, 4),
  mean = c(0, 10, -5),
  sd   = c(1, 2, 0.5)
)
pmap(df_params, rnorm)   # data frames are lists — same call
```

Live example — custom loan payment calculator across 3 scenarios

```{r}
# Monthly payment: P * r / (1 - (1+r)^-n)
scenarios <- list(
  principal = c(200000, 350000, 500000),
  rate_yr   = c(0.04,   0.055,  0.07  ),
  years     = c(15,     20,     30    )
)

pmap_dbl(scenarios, function(principal, rate_yr, years) {
  r <- rate_yr / 12
  n <- years * 12
  round(principal * r / (1 - (1 + r)^(-n)), 2)
})
```

Tilde shorthand with N inputs uses ..1, ..2, ..3

```{r}
pmap_dbl(scenarios,
  ~ {
    r <- ..2 / 12          # ..2 = rate_yr
    n <- ..3 * 12          # ..3 = years
    round(..1 * r / (1 - (1 + r)^(-n)), 2)
  }
)
# For clarity, prefer explicit function(principal, rate_yr, years)
# over ..1 ..2 ..3 — names self-document
```

### `pmap_dbl` vs `pmap` — same rule as `map2`

```r
pmap(params, rnorm)       # list — elements can be any type/length
pmap_dbl(params, fn)      # errors if fn doesn't return exactly 1 double
pmap_chr(params, fn)      # errors if fn doesn't return exactly 1 string
pmap_df(params, fn)       # binds rows into a data frame
```

Here's the conceptual thread that ties all three together — they form a progression:

#### The input-count progression

Think of `map`, `map2`, and `pmap` as the same machine with progressively more input conveyors:

```r
map(.x, fn)             →  one conveyor belt
map2(.x, .y, fn)        →  two parallel conveyor belts
pmap(list(.x,.y,.z), fn) →  N parallel conveyor belts
```

In every case, the function is called once per "row" across all inputs simultaneously. The output shape is determined by the suffix: no suffix = list, `_dbl` = double vector, `_chr` = character vector, `_df` = data frame via `bind_rows`.

#### `map_df()` — the ETL workhorse

The key design insight: your function returns a one-row `data.frame` per element, and `map_df` automatically stacks all those rows into a single data frame. This makes it ideal for the extract-transform-load pattern inside R — read a list of files, parse each one into a structured row, get one clean table back.

The `.id` argument is particularly useful: it names the output column that records which input element produced each row. Without it you lose the provenance of each row.

#### `map2(`) — when your function has two arguments that vary together

The important constraint: `.x` and `.y` must be the same length — they're zipped, not crossed. If you want a cross product (every combination), you'd use `expand.grid()` first, then `pmap`.

```r
# Zipped (what map2 does):    (1,a), (2,b), (3,c)
# Cross product (expand.grid): (1,a), (1,b), (1,c), (2,a), (2,b) ...
```

#### `pmap(`) — the general case, and the data frame trick

The most powerful pattern: because a data frame *is* a list of equal-length columns, you can pass a data frame of parameters directly to `pmap`. Each row becomes one function call. This makes `pmap` the natural tool for anything that looks like a parameter sweep, config-driven processing, or row-wise model fitting — patterns you'd otherwise reach for `rowwise()` + `dplyr::mutate()` for.

```r
# These are identical:
pmap(list(a = 1:3, b = 4:6), fn)
pmap(data.frame(a = 1:3, b = 4:6), fn)
```

The named list/data frame approach is also self-documenting — column names become the argument names inside the function, which eliminates the cryptic `..1`, `..2` indexing of tilde shorthand for anything with more than two inputs.

