---
title: Loop Functions
format: html
---

Writing `for` and `while` loops is useful when programming but not particularly easy when working interactively on the command line. There are some functions which implement looping in an easier way.

- `lapply`: Loops over a list and evaluate a function on each element
- `sapply`: Same as `lapply` but attempts to simplify the result
- `apply`: Applies a function over the margins of an array
- `tapply`: Applies a function over subsets of a vector
- `mapply`: Multivariate version of `lapply`

`split`, an auxiliary function, is also useful, particularly in conjunction with `lapply`.

## `lapply`

`lapply` take three arguments: (1) a list `X`, (2) a function (or the name of a function) `FUN`, and (3) other arguments via its `...` argument. If `X` is not a list, it will be coerced to a list via `as.list`.

```{r}
lapply
```

The actual looping is done internally in C code.

`lapply` always returns a list, regardless of the class of the input.

In the following example, we iterate over the list `x` and apply the `mean` function to each of its two members. `lapply()` returns a new list with the mean of each of the items of `x`.

```{r}
x <- list(a = 1.5, b = rnorm(10))
lapply(x, mean)
```

Similarly, we apply the `mean` function to a list `y`, which contains four elements.

```{r}
y <- list(a = 1.4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(y, mean)
```

In the next example, we pass `runif` to `lapply`. `runif` creates random uniform variables using random number generators.

```{r}
z <- 1:4
lapply(z, runif)
```

Here, I specify a range for numbers generated by `runif` to be created.

```{r}
lapply(z, runif, min = 0, max = 10)
```

`lapply` and associated functions make heavy use of *anonymous* functions.

```{r}
s <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
s

lapply(s, function(m) m[,1])
```

## `sapply`

`sapply` is a variant of  `lapply`. All it does is attempt to simplify the result of `lapply`, if possible.

- If the result is a list where every element is of length `1`, then it returs a vector
- If the result is a list where every element is a vector of the same length (although > 1), then it returns a matrix
- If it can't figure things out, then it returns a list

Calling `lapply(y, mean)` before, returned a list  of length `4`, where every element of the list is a single number.

Calling, `sapply()` instead, returns a vector with four numbers.

```{r}
sapply(y, mean)
```

## `apply`

The `apply()` function is used to evaluate a function (often an anonymous one) over the margins of an array.

- It is most often used to apply a function to the rows or columns of a matrix
- It can be used with general arrays, e.g., taking the average of an array of matrices
- It is not really faster than writing a loop, but it works in one line!

**Syntax**

```{r}
str(apply)
```

Where:

- `X` is an array
- `MARGIN` is an integer vector indicating which margins should be retained
- `FUN` is a function to be applied
- `...` is for other arguments to be passed to `FUN`

So, let's create a 20x10 matrix, containing normal random variables:

```{r}
t <- matrix(rnorm(200), 20, 10)
```

Now, let's use `apply` to calculate the mean of each column of thr matrix:

```{r}
apply(t, 2, mean)
```

In this case, `apply` returns a vector of length `10` containing the mean of each column of the matrix.

Similarly, let's calculate the sum of all the rows:

```{r}
apply(t, 1, sum)
```

### col/row Sums and Means

For simple operations, like calculating the sum, or calculating the mean or the sum of a column or a row, there are special functions that are highly optimized to do this very quickly.

- `rowSums` = `apply(x, 1, sum)`
- `rowMeans` = `apply(x, 1, mean)`
- `colSums` = `apply(x, 2, sum)`
- `colMeans` = `apply(x, 2, mean)`

These shorcuts are *much* faster, but one only notice when working with large matrices.

### Other Ways to Apply

We can use `apply` with other types of functions.

For example, let's say we have a 20x10 matrix, containing random normal veriables.

```{r}
u <- matrix(rnorm(200), 20, 10)
```

Now, let's say we want to calculate the 25th and 75th percentile of each row.

```{r}
apply(u, 1, quantile, probs = c(0.25, 0.75))
```

Notice how we use the `...` argument of the `apply` function to pass the `probs` argument to the `quantile` function.

`apply` creates a matrix that has two rows, and the number of columns is equal to the number of rows in `u`, which happens to be 20. So, `apply` returns a 2x20 matrix, where in each column, we have 25th and 75th percentile for the corresponding row.

**Example: Average matrix in an array**

```{r}
v <- array(rnorm(2 * 2 * 10), c(2, 2, 10))

apply(v, c(1, 2), mean)
```

## `mapply`

`mapply` is a multivariate version of `apply` which applies a function in parallel over a set of arguments.

**Syntax**

```{r}
str(mapply)
```

Where:

- `FUN` is a function to apply
- `...` contains arguments to apply over
- `MoreArgs` is list of other arguments for `FUN`
- `SIMPLIFY` indicates whether the result should be simplified

The following would be tedious to type:

```{r}
list(rep(1, 4), rep(2, 3), rep(3, 2), rep(4, 1))
```

Instead, we could do:

```{r}
mapply(rep, 1:4, 4:1)
```

**Example: Vectorizing a function**

With `mapply` we could vectorize a function that doesn't work with vectors.

```{r}
noise <- function(n, mean, sd) rnorm(n, mean, sd)
```

So calling `noise` with a single set of arguments, we get 5 random numbers with a mean of 1 and a standard deviation of 2.

```{r}
noise(5, 1, 2)
```

However, if we pass vectors, we don't get the expected behavior, i.e., one random normal with mean one, two random normals with mean 2, etc.

```{r}
noise(1:5, 1:5, 2)
```


Using `mapply` however, gets us the expected result.

```{r}
mapply(noise, 1:5, 1:5, 2)
```

## `tapply`

`tapply` is used to apply a function over the subsets of a vector.

**Syntax**

```{r}
str(tapply)
```

Where:

- `X` is a vector
- `INDEX` is a factor or a list of factors (otherwise, they are coerced to factors)
- `FUN` is a function to be applied
- `...` contains others args to be passed to `FUN`
- `simplify` indicates whether to simplify the result

**Example**

Let's simulate a vector of ten normal random variables, ten uniform random variables, and ten normals with a mean of 1.

```{r}
w <- c(rnorm(10), runif(10), rnorm(10, 1))
```

We can think of vector `w` as having three groups.

Now, let's create a factor variable `f`, with three levels repeated 10 times:

```{r}
f <- gl(3, 10)
```

So, by printing `f`, we can see ten ones, ten twos, and ten threes, indicating in which group each observation is in.

Then, we can use `tapply` to get the mean of each group of numbers in `w`.

```{r}
tapply(w, f, mean)
```
