---
title: Loops and Functions in R
format: html
---

In R, loops exist at two different abstraction levels:

1. Imperative loops — you explicitly control iteration (`for`, `while`, `repeat`)
2. Functional iteration — you pass functions into higher-order functions (`apply`, `lapply`, `purrr::map`, etc.)

The second style is heavily preferred in idiomatic R because:

- R is vectorized
- Functional iteration is usually shorter and safer
- It avoid manual indexing bugs
- Many operations are implemented internally with optimized C code

Think of it this way:

- `for` loops = manually driving a car
- `apply` family = using autopilot for common iteration patterns

---

## Basic Loops in R

### `for` Loop

Used when you know how many iterations you need.

```{r}
numbers <- c(1, 2, 3, 4, 5)

for (n in numbers) print(n * 2)
```

#### Internals

R iterates through each element in the vector and binds it to n.

Equivalent mental model:

```{r}
n <- numbers[1]
n <- numbers[2]
```

### Index-based `for`

Useful when position matters.

```{r}
numbers <- c(10, 20, 30)

for (i in seq_along(numbers)) {
    cat("Index: ", i, "Value: ", numbers[i], "\n")
}
```

Use `seq_along()` instead of `1:length(x)` because it safely handles empty vectors.

### `while` loop

Runs while a condition is true.

```{r}
x <- 1

while (x <= 5) {
    print(x)
    x <- x + 1
}
```

Good for:

- unknown iteration counts
- polling
- simulations
- convergence algorithms

### `repeat` loop

Infinite loop until explicitly stopped with `break`.

```{r}
y <- 1

repeat {
    print(y)

    if (y == 5) break

    y <- y + 1
}
```

Useful for:

- retry logic
- event processing
- iterative convergence

## Control Statements

### `break`

Stops a loop entirely.

```{r}
for (i in 1:10) {
    if (i == 5) break
    print(i)
}
```

### `next`

Skips current iteration.

```{r}
for (i in 1:5) {
    if (i  == 3) next
    print(i)
}
```

## Vectorization (Preferred Over Many Loops)

R is fundamentally vector-oriented.

Instead of:

```{r}
result <- c()

for (i in 1:5) result[i] <- i * 2
```

Prefer:

```{r}
result <- 1:5 * 2
```

This is:

- cleaner
- faster
- memory efficient

## The `apply` Family

The `apply` family abstracts iteration patterns.

### `apply()`

Applies a function across rows or columns of matrices/data frames.

Syntax:

```{r}
str(apply)
```

Where:

- `X` = matrix or array
- `MARGIN`
    - `1` = rows
    - `2` = columns
- `FUN` = function
- `...` = other args to pass to `FUN`
- `simplify` = whether the simplify the result

**Example: Column sums**

```{r}
mat <- matrix(1:9, nrow = 3)
apply(mat, 2, sum)
```

**Example: Row means**

```{r}
apply(mat, 1, mean)
```

#### Important Limitation

`apply()` converts data frames into matrices.

This can cause type coercion.

Example:

```{r}
df <- data.frame(
    a = 1:3,
    b = c("x", "y", "z")
)

apply(df, 2, class)
```

Everything becomes characters.

Because of this:

- `apply()` is best for numeric matrices
- use `lapply()` for general lists/data frames

### `lapply()`

Applies a function to each element of a list and returns a list.

Syntax:

```{r}
str(lapply)
```

`lapply` take three arguments: (1) a list `X`, (2) a function (or the name of a function) `FUN`, and (3) other arguments via its `...` argument. If `X` is not a list, it will be coerced to a list via `as.list`.

**Example:**

```{r}
numbers <- list(a = 1:3, b = 4:6)

lapply(numbers, sum)
```

**Key Characteristic**

Always returns a list. This predictability is extremely important in robust programming.

### `sapply()`

Simplified `lapply()`.

Attempts to simplify the result of `lapply()`.

```{r}
sapply(numbers, sum)
```

Instead of a list, it returned a vector.

#### Why `sapply()` Can Be Dangerous

Return types are inconsistent. Depending on data:

- vector
- matrix
- list

This unpredictability can break production code. For reliable code:

- prefer `lapply()`
- or `vapply()`

### `vapply()`

Safer version of `sapply()`.

You explicitly define the output type.

**Syntax:**

```{r}
str(vapply)
```

**Example:**

```{r}
vapply(numbers, sum, numeric(1))
```

This guarantees:

- each result must be numeric
- length 1

If not, R throws an error. This is very valuable in production systems.

### `mapply()`

Multivariate apply. Applies a function to multiple inputs simultaneously. Equivalent to parallel iteration.

Syntax:

```{r}
str(mapply)
```

**Example:**

```{r}
mapply(sum, 1:3, 4:6)
```

Equivalent conceptual loop:

```{r}
sum(1, 4)
sum(2, 5)
sum(3, 6)
```

**Another Example:**

```{r}
mapply(
    function(x, y) x * y,
    c(1, 2, 3),
    c(10, 20, 30)
)
```

### `Map()`

Wrapper around `mapply()` that always returns a list.

```{r}
Map(
    function(x, y) x * y,
    c(1, 2),
    c(10, 20)
)
```

### `tapply()`

Grouped apply. Applies a function over subsets defined by factors.

Syntax:

```{r}
str(tapply)
```

Where:

* `X` = data
* `INDEX` = grouping variable
* `FUN` = function to apply

**Example:**

```{r}
scores <- c(90, 85, 88, 70, 95)
groups <- c("A", "A", "B", "B", "B")

tapply(scores, groups, mean)
```

Equivalent SQL mental model:

```sql
GROUP BY groups
AVG(scores)
```

### `by()`

Object-oriented version of `tapply()`.

Used commonly with data frames.

```{r}
by(
    iris$Sepal.Length,
    iris$Species,
    mean
)
```

### `aggregate()`

Very important for grouped operations on data frames.

```{r}
aggregate(mpg ~ cyl, data = mtcars, mean)
```

Equivalent idea:

```sql
SELECT cyl, AVG(mpg)
GROUP BY cyl
```

### `split()`

Splits data into groups.

```{r}
split(mtcars$mpg, mtcars$cyl)
```

Returns a list grouped by cylinder count.

#### Powerful Pattern

```{r}
groups <- split(mtcars$mpg, mtcars$cyl)
lapply(groups, mean)
```

This is effectively what `tapply()` does internally.

## Modern Functional Iteration

From the `tidyverse` ecosystem.

Library:

```{r}
library(purrr)
```

### `map()`

Equivalent to `lapply()`.

```{r}
map(list(1:3, 4:6), sum)
```

### Typed Maps

Safer and cleaner.

```{r}
map_dbl(list(1:3, 4:6), sum)
```

Returns numeric vector.

Other variants:

* `map_chr()`
* `map_int()`
* `map_lgl()`

### `map2()`

Equivalent to `mapply()`.

```{r}
map2_dbl(c(1, 2), c(10, 20), ~ .x + .y)
```

### `pmap()`

Generalized multi-input iteration. Useful for rows of data frames.

## Performance Considerations

Historically:

* loops in R were slow
* vectorization was dramatically faster

Modern R has improved significantly.

Today:

* preallocated for loops are often perfectly acceptable
* badly written loops remain slow

**Bad Loop**

```{r}
x <- c()

for (i in 1:10000) {
    x <- c(x, i)
}
```

Problem:

* reallocates memory every iteration

**Good Loop**

```{r}
x <- numeric(10000)

for (i in 1:10000) {
    x[i] <- i
}
```

This preallocates memory once. Major performance improvement.

## Mental Model Summary

| Function    | Purpose                   | Returns        |
|-------------|---------------------------|----------------|
| for         | Explicit iteration        | Anything       |
| while       | Conditional looping       | Anything       |
| repeat      | Infinite loop until break | Anything       |
| apply       | Matrix row/column ops     | Vector/list    |
| lapply      | List iteration            | List           |
| sapply      | Simplified list iteration | Simplified     |
| vapply      | Typed/safe apply          | Fixed type     |
| mapply      | Multi-input iteration     | Simplified     |
| Map         | Multi-input iteration     | List           |
| tapply      | Grouped computation       | Array/list     |
| split       | Divide into groups        | List           |
| aggregate   | Grouped summaries         | Data frame     |
| map (purrr) | Functional iteration      | Typed variants |

## Architectural Perspective

In real-world R systems:

* Use vectorization first
* Use apply family second
* Use explicit loops when:
    * state changes between iterations
    * simulations
    * recursive algorithms
    * iterative optimization
    * sequential dependencies exist

Production-grade R code often favors:

```r
vapply()
purrr::map_*()
data.table
dplyr
```

because:

* types are predictable
* pipelines remain composable
* debugging becomes easier
* behavior is more deterministic

## Example: Data Pipeline Style

Traditional loop:

```{r}
results <- numeric(length(groups))

for (i in seq_along(groups)) {
    results[i] <- mean(groups[[i]])
}
```

Functional style:

```{r}
sapply(groups, mean)
```

Modern tidyverse:

```{r}
map_dbl(groups, mean)
```

Same operation. Different abstraction levels.

